Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-06-11 paper.qmd
Updated -0.1-0 range mathematical notations
Updated word colour to color

nldr_two_curvy_original <- ggplot(tsne_two_curvy_scaled, 
                                  aes(x = emb1, y = emb2)) +
  geom_point(alpha = 0.5, color = clr_choice, size = 0.5)

@@

hex_grid_two_curvy <- ggplot(
  data = hex_grid_with_counts, 
  aes(x = x, y = y)) +
  geom_polygon(color = "grey70", 
               aes(group = hex_poly_id), 
               fill = "#ffffff") +
  geom_point(data = tsne_two_curvy_scaled, 
             aes(x = emb1, y = emb2), 
             alpha = 0.5, size = 0.5, color = clr_choice) +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b", sc_ltr_pos, cex = 2)

hex_grid_coloured_two_curvy1 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts, 
    aes(x = x, y = y,
      group = hex_poly_id, 
      fill = std_counts), color = "grey70", linewidth=0.2) +
  geom_point(data = tsne_two_curvy_scaled,
             aes(x = emb1, y = emb2),
             colour = clr_choice,
             alpha = 0.3,
             size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "E") +
  interior_annotation("b", sc_ltr_pos, cex = 2) + 
  xlim(sc_xlims) + ylim(sc_ylims) 
```

@@

hex_grid_nonempty_two_curvy <- ggplot(
  data = hex_grid_nonempty, 
  aes(x = x, y = y)) +
  geom_polygon(color = "grey70", 
               aes(group = hex_poly_id), 
               fill = "#ffffff") +
  geom_point(data = df_bin_centroids_two_curvy, 
             aes(x = c_x, y = c_y), 
             color = "#000000",
             size = 1) +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("c", sc_ltr_pos, cex = 2) 
```

@@

trimesh_two_curvy <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  geom_point(data = tsne_two_curvy_scaled,
             aes(
               x = emb1,
               y = emb2
             ),
             color = "#636363",
             alpha = 0.5,
             size = 0.5
  ) +
  coord_equal() +
  interior_annotation("a1", sc_ltr_pos)

@@

trimesh_two_curvy_removed1_tsne_with_data <- ggplot() +
  geom_point(data = tsne_two_curvy_scaled,
             aes(
               x = emb1,
               y = emb2
             ),
             color = clr_choice,
             alpha = 0.5,
             size = 0.5) +
  geom_segment(data = tr_from_to_df_two_curvy,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  interior_annotation("a", sc_ltr_pos, cex = 1.2) +
  theme(aspect.ratio = 1)
```

@@

                 color = "#bdbdbd") +

@@

      color = "#31a354", size = 0.9) +

@@

               color = "black") + 

@@

              color = "black", 

@@

                 color = "black")+

@@

             label = expression(r[1]), color = "black", size = 8) +

@@

             label = expression(r[2]), color = "black", size = 8) +

@@

        color = "black")+ # a1 = 0.2087578

@@

      color = "black") + # a2 = 0.1807896

@@

             color = "black", hjust=-0.2, size = 8) +

@@

      color = "black", size = 8) +

@@

hex_grid_coloured_two_curvy2 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_two_curvy2, 
    aes(x = x, y = y, 
        group = hex_poly_id, 
        fill = std_counts), 
        color = "grey70", 
        linewidth=0.2) +
  geom_point(data = tsne_two_curvy_scaled,
           aes(x = emb1, y = emb2),
           alpha = 0.3,
           size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "C") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("a", sc_ltr_pos, cex = 2)

trimesh_two_curvy2 <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy2,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  geom_point(data = tsne_two_curvy_scaled,
             aes(
               x = tSNE1,
               y = tSNE2
             ),
             color = "#636363",
            alpha = 0.5,
            size = 0.5
            ) +
  coord_equal() +
  interior_annotation("a1", sc_ltr_pos)

@@

hex_grid_coloured_two_curvy3 <-  ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_two_curvy3, 
    aes(x = x, y = y, 
        group = hex_poly_id, 
        fill = std_counts), 
        color = "grey70", linewidth=0.2) +
  geom_point(data = tsne_two_curvy_scaled,
           aes(x = emb1, y = emb2),
           alpha = 0.3,
           size = 0.5) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "D") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("c", sc_ltr_pos, cex = 2)
```

@@

mse_two_curvy_b <- ggplot(error_two_curvy_umap, 
                     aes(x = a1, 
                         y = RMSE)) +
  geom_vline(xintercept = 0.09,
           color = "#d95f02", linetype="dashed", 
           linewidth=1) +
  geom_vline(xintercept = 0.05,
             color = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = 0.03, linetype="dotted",
             color = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) + 
  geom_point(size = 1) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_umap$a1, 2)))[c(1, 4, 7, 11, 13, 14, 15, 16)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = "RMSE") +
  interior_annotation("a", position = c(0.9, 0.1), cex = 2) +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 18),
        axis.text = element_text(size = 15))

prop_dens_a1 <- ggplot(error_two_curvy_umap,
                     aes(x = a1,
                         y = prop_comp)) +
  geom_vline(xintercept = 0.09,
           color = "#d95f02", linetype="dashed", 
           linewidth=1) +
  geom_vline(xintercept = 0.05,
             color = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = 0.03, linetype="dotted",
             color = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1) +
  # scale_y_continuous(breaks = sort(unique(error_two_curvy$b_non_empty))[-3]) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_umap$a1, 2)))[c(1, 4, 7, 11, 13, 14, 15, 16)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = paste("Relative proportion density")) +
  interior_annotation("d", position = c(0.9, 0.9), cex = 2) +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 18),
axis.text = element_text(size = 15))

@@

a1_m_two_curvy <- ggplot(error_two_curvy_umap,
                     aes(x = a1,
                         y = prop_bins)) +
  geom_vline(xintercept = 0.09,
           color = "#d95f02", linetype="dashed", 
           linewidth=1) +
  geom_vline(xintercept = 0.05,
             color = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = 0.03, linetype="dotted",
             color = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1) +
  # scale_y_continuous(breaks = sort(unique(error_two_curvy$b_non_empty))[-3]) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_umap$a1, 2)))[c(1, 4, 7, 11, 13, 14, 15, 16)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = expression("prop. non-empty bins " * bgroup("(", m / b, ")"))) +
  interior_annotation("b", position = c(0.9, 0.1), cex = 2) +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 18),
axis.text = element_text(size = 15))
```

```{r}
#| label: fig-p-d-error-in-2d-two-curvy
#| fig-cap: "The $4\\text{-}D$ model error in \\gD{} layout. Color indicates error ($e_{hj}$), dark colour indicating high error and light indicates low error. Most large errors are distributed near the sparse end of the non-linear cluster with dense corner."
#| out-width: 80%
#| fig-pos: H

@@

The values $(s_1, \ s_2)$ define the position of the centroid of the bottom left hexagon. By default, this is at $s_1 = -q, s_2 = -qr_2$, where $q$ is the buffer bound the data. The choice of these values can have some effect on the distribution of bin counts. @fig-bins-two-curvy illustrates this. The distribution of bin counts for $s_1$ varying between $-0.1-0.0$ is shown. Generally, a more uniform distribution among these possibilities would indicate that the bins are reliably capturing the underlying distribution of observations.

@@

mse_two_curvy_lwd <- ggplot(error_rm_two_curvy, 
                     aes(x = benchmark_rm_lwd, 
                         y = RMSE,
                         color = a1,
                         linetype = a1)) + 
  geom_point(
    size = 1
    ) +
  geom_line(
    linewidth = 0.5
    ) + 
  scale_x_continuous(breaks = sort(unique(error_rm_two_curvy$benchmark_rm_lwd))[c(1, 3, 5, 7, 9, 11, 13, 15, 17)]) +
  scale_color_manual(values=c("#d95f02", "#1f78b4", "#1b9e77"),
                     labels = parse_format()) +
  scale_linetype_manual(values = c("dashed", "solid", "dotted"),
                        labels = parse_format()) +
  xlab("bin density cutoff") +
  ylab("RMSE") +
  interior_annotation("c", position = c(0.07, 0.9), cex = 2) +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.text = element_text(size = 18),
        axis.title = element_text(size = 18),
axis.text = element_text(size = 15))
```

@@

@fig-toy-mse illustrates the approach to compare the fits for different representations and assess the strength of any fit. What does it mean to be a best fit for this problem? Analysts use an NLDR layout to display the structure present in high-dimensional data in a convenient 2D display. It is a competitor to linear dimension reduction that can better represent non-linear associations such as clusters. However, these methods can hallucinate, suggesting patterns that don't exist, and grossly exaggerate other patterns. Having a layout that best fits the high-dimensional structure is desirable but more important is to identify bad representations so they can be avoided. The goal is to help users decide on a the most useful and appropriate low-dimensional representation of the high-dimensional data.

A particular pattern that we commonly see is that analysts tend to pick layouts with clusters that have big separations between them. When you examine their data in a tour, it is almost always that we see there are no big separations, and actually the suggested clusters are not even present. While we don't expect that analysts include animated gifs of tours in their papers, we should expect that any 2D representation adequately indicates the clustering that is present, and honestly show lack of separation or lack of clustering when it doesn't exist. It is important for analysts to have tools to select the accurate representation not the pretty but wrong representation.

@@

```{r}
#| fig-cap: "Assessing which of the 6 NLDR layouts on the two non-linear clusters data is the better representation using RMSE for varying binwidth ($a_1$). Colour used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-f). Layout d is universally poor. Layouts a, b, e that show two close clusters are universally suboptimal. Layout b with little separation performs well at tiny binwidth (where most points are in their own bin) and poorly as binwidth increases. Layout e has small separation with oddly shaped clusters. Layout a is the best choice."
#| label: fig-toy-mse
#| fig-pos: H
#| out-width: 100%
#| fig-height: 6
#| fig-width: 6

@@

trimesh_removed_gau_tsne <- ggplot() + 
  geom_point(
    data = tsne_gau_scaled,
    aes(
      x = tSNE1,
      y = tSNE2
    ),
    colour = clr_choice,
    alpha = 0.1) + 
  geom_segment(
    data = tr_from_to_df_gau1,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to),
    color = "#000000",
    #alpha = 0.4,
    linewidth = 0.5) +
  interior_annotation("a1", 
                      position = c(0.08, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )

@@

five_gau_proj_tsne_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_point(
    #size = 0.5,
    alpha = 0.05,
    color = clr_choice) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000",
    alpha = 0.4,
    linewidth = 0.8) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2),
    label=rownames(axes), 
    colour="grey50",
    size = 5) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.89, 0.89)) +
  ylim(c(-0.89, 0.89)) +
  interior_annotation("a2", 
                      position = c(0.08, 0.9),
                      cex = 2)

@@

trimesh_removed_gau_umap <- ggplot() + 
  geom_point(
    data = umap_gau_scaled,
    aes(
      x = UMAP1,
      y = UMAP2
    ),
    colour = clr_choice,
    alpha = 0.05) + 
  geom_segment(
    data = tr_from_to_df_gau1,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to),
    color = "#000000",
    #alpha = 0.4,
    linewidth = 0.5) +
  interior_annotation("b1", 
                      position = c(0.95, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )

@@

five_gau_proj_umap_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_point(
    alpha = 0.05,
    color = clr_choice) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000",
    alpha = 0.4,
    linewidth = 0.8) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 5) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.89, 0.89)) +
  ylim(c(-0.89, 0.89)) +
  interior_annotation("b2", 
                      position = c(0.08, 0.9),
                      cex = 2)

@@

trimesh_removed_gau_pacmap <- ggplot() + 
  geom_point(
    data = pacmap_gau_scaled,
    aes(
      x = PaCMAP1,
      y = PaCMAP2),
      colour = clr_choice,
    alpha = 0.1) +
  geom_segment(
    data = tr_from_to_df_gau1,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to),
    color = "#000000",
    #alpha = 0.4,
    linewidth = 0.5) +
  interior_annotation("c1", 
                      position = c(0.08, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )

@@

error_plot_one_curvy_hist_selected <- ggplot(error_df_one_curvy_abs_deselected11, 
      aes(x = sqrt_row_wise_total_error, 
          y = density)) +
  geom_point(alpha=0.7, color = "#d9d9d9") +
  geom_point(data = error_df_one_curvy_abs_selected11, 
      aes(x = sqrt_row_wise_total_error, 
          y = density),
      alpha=0.7, color = "#800026", size = 3) +
  xlab(expression(e[hj])) +
  ylab("") +
  interior_annotation("b1",
                      cex = 2) +
  theme_bw() +
  theme(aspect.ratio = 1,
        axis.title = element_text(size = 18),
        axis.text = element_text(size = 15))

>>>>

nldr_two_curvy_original <- ggplot(tsne_two_curvy_scaled, 
                                  aes(x = emb1, y = emb2)) +
  geom_point(alpha = 0.5, colour = clr_choice, size = 0.5)

@@

hex_grid_two_curvy <- ggplot(
  data = hex_grid_with_counts, 
  aes(x = x, y = y)) +
  geom_polygon(colour = "grey70", 
               aes(group = hex_poly_id), 
               fill = "#ffffff") +
  geom_point(data = tsne_two_curvy_scaled, 
             aes(x = emb1, y = emb2), 
             alpha = 0.5, size = 0.5, colour = clr_choice) +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b", sc_ltr_pos, cex = 2)

hex_grid_coloured_two_curvy1 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts, 
    aes(x = x, y = y,
      group = hex_poly_id, 
      fill = std_counts), colour = "grey70", linewidth=0.2) +
  geom_point(data = tsne_two_curvy_scaled,
             aes(x = emb1, y = emb2),
             colour = clr_choice,
             alpha = 0.3,
             size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "E") +
  interior_annotation("b", sc_ltr_pos, cex = 2) + 
  xlim(sc_xlims) + ylim(sc_ylims) 
```

@@

hex_grid_nonempty_two_curvy <- ggplot(
  data = hex_grid_nonempty, 
  aes(x = x, y = y)) +
  geom_polygon(colour = "grey70", 
               aes(group = hex_poly_id), 
               fill = "#ffffff") +
  geom_point(data = df_bin_centroids_two_curvy, 
             aes(x = c_x, y = c_y), 
             colour = "#000000",
             size = 1) +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("c", sc_ltr_pos, cex = 2) 
```

@@

trimesh_two_curvy <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  geom_point(data = tsne_two_curvy_scaled,
             aes(
               x = emb1,
               y = emb2
             ),
             colour = "#636363",
             alpha = 0.5,
             size = 0.5
  ) +
  coord_equal() +
  interior_annotation("a1", sc_ltr_pos)

@@

trimesh_two_curvy_removed1_tsne_with_data <- ggplot() +
  geom_point(data = tsne_two_curvy_scaled,
             aes(
               x = emb1,
               y = emb2
             ),
             colour = clr_choice,
             alpha = 0.5,
             size = 0.5) +
  geom_segment(data = tr_from_to_df_two_curvy,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  interior_annotation("a", sc_ltr_pos, cex = 1.2) +
  theme(aspect.ratio = 1)
```

@@

                 colour = "#bdbdbd") +

@@

      colour = "#31a354", size = 0.9) +

@@

               colour = "black") + 

@@

              colour = "black", 

@@

                 colour = "black")+

@@

             label = expression(r[1]), colour = "black", size = 8) +

@@

             label = expression(r[2]), colour = "black", size = 8) +

@@

        colour = "black")+ # a1 = 0.2087578

@@

      colour = "black") + # a2 = 0.1807896

@@

             colour = "black", hjust=-0.2, size = 8) +

@@

      colour = "black", size = 8) +

@@

hex_grid_coloured_two_curvy2 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_two_curvy2, 
    aes(x = x, y = y, 
        group = hex_poly_id, 
        fill = std_counts), 
        colour = "grey70", 
        linewidth=0.2) +
  geom_point(data = tsne_two_curvy_scaled,
           aes(x = emb1, y = emb2),
           alpha = 0.3,
           size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "C") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("a", sc_ltr_pos, cex = 2)

trimesh_two_curvy2 <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy2,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  geom_point(data = tsne_two_curvy_scaled,
             aes(
               x = tSNE1,
               y = tSNE2
             ),
             colour = "#636363",
            alpha = 0.5,
            size = 0.5
            ) +
  coord_equal() +
  interior_annotation("a1", sc_ltr_pos)

@@

hex_grid_coloured_two_curvy3 <-  ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_two_curvy3, 
    aes(x = x, y = y, 
        group = hex_poly_id, 
        fill = std_counts), 
        colour = "grey70", linewidth=0.2) +
  geom_point(data = tsne_two_curvy_scaled,
           aes(x = emb1, y = emb2),
           alpha = 0.3,
           size = 0.5) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "D") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("c", sc_ltr_pos, cex = 2)
```

@@

mse_two_curvy_b <- ggplot(error_two_curvy_umap, 
                     aes(x = a1, 
                         y = RMSE)) +
  geom_vline(xintercept = 0.09,
           colour = "#d95f02", linetype="dashed", 
           linewidth=1) +
  geom_vline(xintercept = 0.05,
             colour = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = 0.03, linetype="dotted",
             colour = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) + 
  geom_point(size = 1) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_umap$a1, 2)))[c(1, 4, 7, 11, 13, 14, 15, 16)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = "RMSE") +
  interior_annotation("a", position = c(0.9, 0.1), cex = 2) +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 18),
        axis.text = element_text(size = 15))

prop_dens_a1 <- ggplot(error_two_curvy_umap,
                     aes(x = a1,
                         y = prop_comp)) +
  geom_vline(xintercept = 0.09,
           colour = "#d95f02", linetype="dashed", 
           linewidth=1) +
  geom_vline(xintercept = 0.05,
             colour = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = 0.03, linetype="dotted",
             colour = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1) +
  # scale_y_continuous(breaks = sort(unique(error_two_curvy$b_non_empty))[-3]) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_umap$a1, 2)))[c(1, 4, 7, 11, 13, 14, 15, 16)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = paste("Relative proportion density")) +
  interior_annotation("d", position = c(0.9, 0.9), cex = 2) +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 18),
axis.text = element_text(size = 15))

@@

a1_m_two_curvy <- ggplot(error_two_curvy_umap,
                     aes(x = a1,
                         y = prop_bins)) +
  geom_vline(xintercept = 0.09,
           colour = "#d95f02", linetype="dashed", 
           linewidth=1) +
  geom_vline(xintercept = 0.05,
             colour = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = 0.03, linetype="dotted",
             colour = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1) +
  # scale_y_continuous(breaks = sort(unique(error_two_curvy$b_non_empty))[-3]) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_umap$a1, 2)))[c(1, 4, 7, 11, 13, 14, 15, 16)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = expression("prop. non-empty bins " * bgroup("(", m / b, ")"))) +
  interior_annotation("b", position = c(0.9, 0.1), cex = 2) +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 18),
axis.text = element_text(size = 15))
```

```{r}
#| label: fig-p-d-error-in-2d-two-curvy
#| fig-cap: "The $4\\text{-}D$ model error in \\gD{} layout. Color indicates error ($e_{hj}$), dark color indicating high error and light indicates low error. Most large errors are distributed near the sparse end of the non-linear cluster with dense corner."
#| out-width: 80%
#| fig-pos: H

@@

The values $(s_1, \ s_2)$ define the position of the centroid of the bottom left hexagon. By default, this is at $s_1 = \text{-}q, s_2 = \text{-}qr_2$, where $q$ is the buffer bound the data. The choice of these values can have some effect on the distribution of bin counts. @fig-bins-two-curvy illustrates this. The distribution of bin counts for $s_1$ varying between $\text{-}0.1\,\text{--}\,0.0$ is shown. Generally, a more uniform distribution among these possibilities would indicate that the bins are reliably capturing the underlying distribution of observations.

@@

mse_two_curvy_lwd <- ggplot(error_rm_two_curvy, 
                     aes(x = benchmark_rm_lwd, 
                         y = RMSE,
                         colour = a1,
                         linetype = a1)) + 
  geom_point(
    size = 1
    ) +
  geom_line(
    linewidth = 0.5
    ) + 
  scale_x_continuous(breaks = sort(unique(error_rm_two_curvy$benchmark_rm_lwd))[c(1, 3, 5, 7, 9, 11, 13, 15, 17)]) +
  scale_color_manual(values=c("#d95f02", "#1f78b4", "#1b9e77"),
                     labels = parse_format()) +
  scale_linetype_manual(values = c("dashed", "solid", "dotted"),
                        labels = parse_format()) +
  xlab("bin density cutoff") +
  ylab("RMSE") +
  interior_annotation("c", position = c(0.07, 0.9), cex = 2) +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.text = element_text(size = 18),
        axis.title = element_text(size = 18),
axis.text = element_text(size = 15))
```

@@

@fig-toy-mse illustrates the approach to compare the fits for different representations and assess the strength of any fit. What does it mean to be a best fit for this problem? Analysts use an NLDR layout to display the structure present in high-dimensional data in a convenient \gD{} display. It is a competitor to linear dimension reduction that can better represent non-linear associations such as clusters. However, these methods can hallucinate, suggesting patterns that don't exist, and grossly exaggerate other patterns. Having a layout that best fits the high-dimensional structure is desirable but more important is to identify bad representations so they can be avoided. The goal is to help users decide on a the most useful and appropriate low-dimensional representation of the high-dimensional data.

A particular pattern that we commonly see is that analysts tend to pick layouts with clusters that have big separations between them. When you examine their data in a tour, it is almost always that we see there are no big separations, and actually the suggested clusters are not even present. While we don't expect that analysts include animated gifs of tours in their papers, we should expect that any \gD{} representation adequately indicates the clustering that is present, and honestly show lack of separation or lack of clustering when it doesn't exist. It is important for analysts to have tools to select the accurate representation not the pretty but wrong representation.

@@

```{r}
#| fig-cap: "Assessing which of the 6 NLDR layouts on the two non-linear clusters data is the better representation using RMSE for varying binwidth ($a_1$). Color used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-f). Layout d is universally poor. Layouts a, b, e that show two close clusters are universally suboptimal. Layout b with little separation performs well at tiny binwidth (where most points are in their own bin) and poorly as binwidth increases. Layout e has small separation with oddly shaped clusters. Layout a is the best choice."
#| label: fig-toy-mse
#| fig-pos: H
#| out-width: 100%
#| fig-height: 6
#| fig-width: 6

@@

trimesh_removed_gau_tsne <- ggplot() + 
  geom_point(
    data = tsne_gau_scaled,
    aes(
      x = tSNE1,
      y = tSNE2
    ),
    colour = clr_choice,
    alpha = 0.1) + 
  geom_segment(
    data = tr_from_to_df_gau1,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to),
    colour = "#000000",
    #alpha = 0.4,
    linewidth = 0.5) +
  interior_annotation("a1", 
                      position = c(0.08, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )

@@

five_gau_proj_tsne_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_point(
    #size = 0.5,
    alpha = 0.05,
    colour = clr_choice) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    colour = "#000000",
    alpha = 0.4,
    linewidth = 0.8) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2),
    label=rownames(axes), 
    colour="grey50",
    size = 5) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.89, 0.89)) +
  ylim(c(-0.89, 0.89)) +
  interior_annotation("a2", 
                      position = c(0.08, 0.9),
                      cex = 2)

@@

trimesh_removed_gau_umap <- ggplot() + 
  geom_point(
    data = umap_gau_scaled,
    aes(
      x = UMAP1,
      y = UMAP2
    ),
    colour = clr_choice,
    alpha = 0.05) + 
  geom_segment(
    data = tr_from_to_df_gau1,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to),
    colour = "#000000",
    #alpha = 0.4,
    linewidth = 0.5) +
  interior_annotation("b1", 
                      position = c(0.95, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )

@@

five_gau_proj_umap_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_point(
    alpha = 0.05,
    colour = clr_choice) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    colour = "#000000",
    alpha = 0.4,
    linewidth = 0.8) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 5) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.89, 0.89)) +
  ylim(c(-0.89, 0.89)) +
  interior_annotation("b2", 
                      position = c(0.08, 0.9),
                      cex = 2)

@@

trimesh_removed_gau_pacmap <- ggplot() + 
  geom_point(
    data = pacmap_gau_scaled,
    aes(
      x = PaCMAP1,
      y = PaCMAP2),
      colour = clr_choice,
    alpha = 0.1) +
  geom_segment(
    data = tr_from_to_df_gau1,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to),
    colour = "#000000",
    #alpha = 0.4,
    linewidth = 0.5) +
  interior_annotation("c1", 
                      position = c(0.08, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )

@@

error_plot_one_curvy_hist_selected <- ggplot(error_df_one_curvy_abs_deselected11, 
      aes(x = sqrt_row_wise_total_error, 
          y = density)) +
  geom_point(alpha=0.7, colour = "#d9d9d9") +
  geom_point(data = error_df_one_curvy_abs_selected11, 
      aes(x = sqrt_row_wise_total_error, 
          y = density),
      alpha=0.7, colour = "#800026", size = 3) +
  xlab(expression(e[hj])) +
  ylab("") +
  interior_annotation("b1",
                      cex = 2) +
  theme_bw() +
  theme(aspect.ratio = 1,
        axis.title = element_text(size = 18),
        axis.text = element_text(size = 15))
